Introducing ICBe: An Event Extraction Dataset from Narratives about International Crises

### 2023-08-16 replication_paper/psrm_draft/ICBe_manuscript.tex
AG penultimate RR

\author{
    Rex W. Douglass
    \thanks{Correspondence should be addressed to Rex W. Douglass at
\href{mailto:rexdouglass@gmail.com}{\nolinkurl{rexdouglass@gmail.com}}.}
   \\
    University of California, San Diego \\
   \\
  \texttt{} \\
   \And
    Thomas Leo Scherer
   \\
    University of California, San Diego \\
   \\
  \texttt{} \\
   \And
    J. Andrés Gannon
   \\
    Vanderbilt University \\
   \\
  \texttt{} \\
   \And
    Erik Gartzke
   \\
    University of California, San Diego \\
   \\
  \texttt{} \\
   \And
    Jon Lindsay
   \\
    Georgia Institute of Technology \\
   \\
  \texttt{} \\
   \And
    Shannon Carcelli
   \\
    University of Maryland \\
   \\
  \texttt{} \\
   \And
    Jonathan Wilkenfeld
   \\
    University of Maryland \\
   \\
  \texttt{} \\
   \And
    David M. Quinn
   \\
    University of Maryland \\
   \\
  \texttt{} \\
   \And
    Catherine Aiken
   \\
    Georgetown University \\
   \\
  \texttt{} \\
   \And
    Jose Miguel Cabezas Navarro
   \\
    Universidad Mayor \\
   \\
  \texttt{} \\
   \And
    Neil Lund
   \\
    University of Maryland \\
   \\
  \texttt{} \\
   \And
    Egle Murauskaite
   \\
    University of Maryland \\
   \\
  \texttt{} \\
   \And
    Diana Partridge
   \\
    University of Maryland \\
   \\
  \texttt{} \\
  }

@@

Much like the recording of chess evolved from natural language
descriptive text notation to the modern figurine algebra notion,
international relations scholars have recently sought to produce a
structured account of historical events by combining the unstructured
corpus of historical records with informative priors about international
relations. The resulting structured account of information about a
historical episode can be combined with that of other events to produce
a systematic account of political events from which we can garner novel
insight, assuming the underlying data have high coverage, precision, and
recall. The easiest way to convey the desired produced of this task is
with an example. (\textbf{fig-cuba-narrative?}) shows a narrative
account of the Cuban Missile Crisis (1962) in natural language sentences
alongside a mapping to discrete machine-readable abstractive events.
From the structured data, scholars can identify similarities and
differences across events concerning important concepts like when
particular foreign policy actions deter versus inflame (Jervis 1978;
Glaser 2000), when third parties mediate in interstate disputes (Haffar
2002; Quinn et al. 2006), and how actors try to communicate resolve
(Trager 2016; Lupton 2018). Identifying patterns of international
interactions is not just an inherently interesting enterprise; it is a
necessary precondition to important efforts to predict where
policymakers should turn their attention to improve global welfare (Ward
et al. 2013; Beger, Morgan, and Ward 2021).

\begin{figure}
\hypertarget{fig-cuba-narrative}{%
\centering
\includegraphics[width=7in,height=8in]{case_study_cuban_precision.png}
\caption{Cuban Missile Crisis (1962) - ICB Narrative vs.~ICBe
Events}\label{fig-cuba-narrative}
}
\end{figure}

@@

With the exception of large-scale CAMEO dictionary-based systems (the
first grouping of columns), our ontology improves upon the the existing
state of the art quantitative datasets that ignore important information
content about international interactions.\footnote{See Balali, Asadpour,
  and Jafari (2021) for a recent review of ontological depth and
  availability of Gold Standard example text.} We highlight two
particular innovations. First, we separate the `chess pieces' from the
`chess players' in distinguishing between different actors within a
state. By virtue of our ontology coding military versus civilian actors
and national leaders versus bureaucrats, our data can be used to explore
important questions concerning civilian-military relations (Narang and
Talmadge 2017), Track Two diplomacy and the role of sub-national actors
(Hsu et al. 2020), and the evolution of what actors are engaged in
crises - a topic of increasing interest as states engage in gray zone
conflict by employing the coast guard or paramilitary mercenaries
instead of internationally recognized state militaries (J. Andrés Gannon
et al. 2022). Second, we add information about the domains in which
actors behave - whether in land, air, sea, space, or cyber - since they
differ in their technology, tactics, geography, and purpose (Gartzke and
Lindsay 2019). Doing so allows researchers to identify and explain
patterns in escalation conditional on the military means states use in
conflict. Recent concerns about cross-domain conflict and the effect of
new domains of conflict like space and cyber have made this an endeavor
of increased interest to practitioners (J. Andrés Gannon 2022).

@@

The ICBe ontology follows a hierarchical design philosophy where a
smaller number of significant decisions are made early on and then
progressively refined into more specific details (Brust and Denzler
2020).\footnote{This process quickly focuses the coder on a smaller
  number of relevant options while also allowing them to apply multiple
  tags if the sentence explicitly includes more than one or there is
  insufficient evidence to choose only one tag. The guided coding
  process also allows for the possibility that earlier coarse decisions
  have less error than later fine-grained decisions.} Each coder was
instructed to first thoroughly read the full crisis narrative and then
presented with a custom graphical user interface (GUI) (SI Appendix
2.1). Coders then proceeded sentence by sentence, choosing the number of
events (0-3) that occurred, the highest behavior (thought, speech, or
activity), a set of players, whether the means were primarily armed or
unarmed, whether there was an increase or decrease in aggression
(uncooperative/escalating or cooperative/de-escalating), and finally one
or more specific and non-mutually exclusive activities. Some additional
details were always collected (e.g.~location and timing) while other
details were only collected if appropriate (e.g.~force size, fatalities,
domains, units). We find 472 actors and 117 different
behaviors.\footnote{See the full codebook on Github Repository
  \href{https://urldefense.com/v3/__https://github.com/CenterForPeaceAndSecurityStudies/ICBEventData__;!!Mih3wA!WxDJtEczKfxGTh0S2Krunap8ReymFEL5iTWaSfOHeqlSdyfRx77zmjBSWO1OAm13$}{ICBEventData}.}
A unique feature of the ontology is that thought, speech, and do
behaviors can be nested into combinations, e.g.~an offer for the
U.S.S.R. to remove missiles from Cuba in exchange for the U.S. removing
missiles from Turkey. Through compounding, the ontology can capture what
players were said to have known, learned, or said about other specific
fully described actions.

@@

Our strategy for dealing with both problems is a plausibly objective
yardstick called a synthetic historical narrative. We collect a large
diverse corpus of narratives spanning timelines, encyclopedia entries,
journal articles, news reports, websites, and government documents.
Using natural language processing (fully described in SI Appendix 3.1),
we identify details that appear across multiple accounts. The more
accounts that mention a detail, the more central it is to understanding
the true historical episode. The theoretical motivation is that authors
face word limits which force them to pick and choose which details to
include, and they choose details that serve the specific context of the
document they are producing. With a sufficiently large and diverse
corpus of documents, we can vary the context while holding the overall
episode constant and see which details tend to be invariant to context.
Intuitively, a high-quality event dataset should have high recall for
context invariant details both because of their broader relevance and
also because they are easier to find in source material.

We find substantive variation in recall across existing state of the art
methods. Mentions of a detail across accounts are exponentially
distributed with context-invariant details appearing dozens to hundreds
of times more than context-dependent details.\footnote{As the ICB
  narratives are intended to explain conflictual behavior in a political
  context, many of the missing events concern more economic components
  of conflict (eg. nationalizing a foreign business). Even when they
  occur in the context of a crisis, these events largely fall outside
  the sample of information on which ICBe's ontology is currently
  trained. Even with this limitation, ICBe is more comprehensive than
  the existing datasets that do try to code the economic dimensions of
  these crises. We see expanding the ontology to broader international
  phenomenon as a promising future implementation of our model.}
Furthermore, crisis start and stop dates are arbitrary, and the
historical record points to many precursor events as necessary detail
for understanding later events. (\textbf{fig-recall-cases?}) compares
ICBe's recall with that of existing datasets for the two case studies
detailed in Section 4. ICBe strictly dominates all of the systems but
ICEWs in recall though we note that the small sample sizes mean these
systems should be considered statistically indistinguishable. Across all
existing datasets and ICBe, recall increases with the number of document
mentions which is an important sign of validity for both them and our
benchmark. The one outlier is Phoenix which in the Cuban Missile Crisis
case is so noisy that it's recall curve is flat to decreasing as
mentions increase. The two episode-level datasets (MIDs and ICM) have
low coverage of contextual details. The two other dictionary systems
ICEWs and Terrier have higher coverage, with ICEWs outperforming
Terrier. Importantly our corpus of ICB narratives has high recall of
frequently mentioned details giving us confidence in how those summaries
were constructed, and ICBe lags only slightly behind showing that it
left little additional information on the table.

@@

How does ICBe's precision compare to the existing state of the art? A
researcher should be able to lay out the events of a crisis on a
timeline and read off the macrostructure of an episode from each
individual move. We call this visualization a crisis map, a directed
graph intersected with a timeline. A crisis map using ICBe for the Cuban
Missile Crisis case study is provided in (\textbf{fig-cuba-crisismap?}),
and crisis maps for the two case studies using existing event datasets
can be found in SI Appendix 4.3 and 4.4 and crisis maps for all crises
using all datasets can be found on the companion website. The crisis
maps reveal the episode-level datasets like MIDs or the original ICB are
too sparse and vague to reconstruct the structure of the crisis (SI
Appendix 4.3 and 4.4). On the other end of the spectrum, the high recall
dictionary-based event datasets like Terrier and ICEWs produce so many
noisy events (several hundred thousand) that even with heavy filtering
their crisis maps are completely unintelligible. Further, because of
copyright issues, none of these datasets directly provide the original
text spans making event-level precision difficult to verify.

@@

(\textbf{fig-umap?}) shows the location of every sentence from the ICBe
corpus in semantic space as embedded using the same large language model
as before, and the median location of each ICBe event tag applied to
those sentences.\footnote{We preprocess sentences to replace named
  entities with a generic Entity token.} Labels reflect the individual
leaves of the ontology and colors reflect the higher level coarse branch
nodes of the ontology. If ICBe has high precision, substantively similar
tags ought to have been applied to substantively similar source text,
which is what we see both in two dimensions in the main plot and via
hierarchical clustering on all dimensions in the dendrogram along the
right-hand side.\footnote{Hierarchical clustering on cosine similarity
  and with Ward's method.}

@@

In this section, we focus our validation on two case studies for which
we have produced synthetic narratives using the method described in
Section 3.2. Our proposed measure is a reconstruction task to see
whether our intended ontology can be recovered through only unsupervised
clustering of sentences they were applied to. The first
((\textbf{fig-cuba-narrative?})) is the Cuban Missile Crisis (hereafter
Cuban Missiles) which took place primarily in the second half of 1962,
involved the United States, the Soviet Union, and Cuba, and is widely
known for bringing the world to the brink of nuclear war. The second (SI
Appendix 4.1) is the Crimea-Donbas Crisis (hereafter Crimea-Donbas)
which took place primarily in 2014, involved Russia, Ukraine, and NATO,
and within a decade spiraled into a full-scale invasion. We choose these
cases because they are significant in contemporary international
relations, are widely known across academic disciplines as well as among
the public, and are sufficiently brief to evaluate in depth. They are
similar in that both cases involve a superpower in crisis with a
neighbor that changed from a friendly to a hostile regime, both held
implications for the economic and military security for the superpower
by risking full-scale invasion, and both eventually invited intervention
by an opposing superpower.

@@

A synthetic historical narrative for Cuban Missiles appears in
(\textbf{fig-recall-cuba?}), with 51 events drawn from 2,020 documents.
Each row represents a detail that appeared in at least five documents
along with an approximate start date, a handwritten summary, the number
of documents it was mentioned in, and whether it could be identified in
the text of the original ICB corpus, our ICBe events, and any of the
competing existing models.

ICB's improved recall of Cuban Missiles was relative to the state of the
art was summarized in Section 3.2 ((\textbf{fig-recall-cases?})), but
the events that explain that improvement can now be seen. Our ground
truth ICB narrative contains 17/51 of the events from the synthetic
narrative of a case that includes high-level previously classified
details. ICBe captures nearly all details included in ICB as well as
more details from the synthetic narrative than any competing dataset.
Phoenix includes some earlier information than ICBe like the
nationalization of businesses and back channel negotiations, but the
crisis narrative has a clean canonical end with the Soviets agreeing to
withdraw missiles. ICBe stands out in including more communicative
behavior (do -- speech) than existing datasets like US threats to attack
and later promises not to invade. Given the recognized importance of
threat credibility for understanding international conflict, the
addition of this information is a substantively important improvement
over the existing state of the art (Slantchev 2011).

@@

(\textbf{fig-cuba-crisismap?}) shows the crisis map for the Cuban
Missile Crisis. Looking at the crisis on a timeline, one can now
identify the structure of actors and the environment, along with its
supporting details, in a way that validates the precision of ICBe.
Although harder to measure objectively, this crisis map provides face
valid evidence that ICBe's account is not too vague, but also not
unnecessarily detailed. We include much of the geopolitically important
details like Soviet deployment, US discovery of that deployment,
heightened alert levels, a blockade, and negotiations that ended with a
formal agreement. At the same time, the crisis map indicates that ICBe
does not include unnecessary nuances that preclude useful comparison to
other international events.

@@

A synthetic historical narrative for the 2014 Crimea-Donbas crisis (30
events drawn from 971 documents) appears in
(\textbf{fig-recall-crimea?}). As in the earlier case, rows represent
details that appeared in at least five documents and whether it is
identified in ICBe and existing datasets.

Again quantitatively summarized earlier in Section 3.2
((\textbf{fig-recall-cases?})), our ground truth ICB narrative contains
23/30 of the events from the synthetic narrative. Like the gray zone
precursor to the Cuban Missile crisis (Cormac and Aldrich 2018), Ukraine
provided several security guarantees to Russia that were potentially
undone, e.g.~a long term lease on naval facilities in Crimea. But unlike
the Cuban Missile crisis, the end of this crisis is unclear, with the
event meekly ending with a second cease-fire agreement (Minsk II) but
continued fighting. ICBe again recalls more important information about
the crisis than any existing dataset, particularly information
concerning the behavior of non-state separatist groups like the Donetsk
People's Republic (DPR) and Luhansk People's Republic (LPR).

@@

We take the same approach here in (\textbf{fig-precision-icews?}),
selecting four specific CAMEO event codings and checking how often they
reflect a true real-world event from the Crimea-Donbas synthetic
narrative. We choose four event types around key moments in the crisis.
The start of the crisis revolves around Ukraine backing out of a trade
deal with the EU in favor of Russia, but ``sign formal agreement''
events act more like a topic detector with dozens of events generated by
discussions of a possible agreement but not the actual agreement which
never materialized. The switch is caught by the ``reject plan, agreement
to settle dispute'', but also continues for Viktor Yanukovych even after
he was removed from power because of articles retroactively discussing
the cause of his removal. Events for ``use conventional military force''
capture a threshold around the start of hostilities and who the
participants were but not any particular battles or campaigns. Likewise,
``impose embargo, boycott, or sanctions'' captures the start of waves of
sanctions and from who but are effectively constant as the news coverage
does not distinguish between subtle changes or additions. In sum,
dictionary-based methods on news corpora tend to have high recall
because they parse everything in the news, but for the same reason,
their specificity for most event types is too low to back out individual
chess-like sequencing that ICBe aims to record.

@@

We release several open-source products along with supporting code and
documentation to further advance the study of international relations,
event extraction, and natural language processing. The first is the
International Crisis Behavior Events (ICBe) dataset, an event-level
aggregation of what took place during the crises identified by the ICB
project. These data are appropriate for statistical analysis of hard
questions about the sequencing of events (e.g.~escalation and
de-escalation of conflicts).\footnote{Using ICBe data J. Andrés Gannon
  (2022) finds that cross-domain crises are shorter and less violent
  than same-domain crises.} Second, we provide a coder-level
disaggregation with multiple codings of each sentence by experts and
undergrads that allows for the introduction of uncertainty and human
interpretation of events. Further, we release a direct mapping from the
codings to the source text at the sentence level as a new resource for
natural language processing. Finally, we provide a companion website
that incorporates detailed visualizations of all of the data introduced
here at crisisevents.org.

@@

\leavevmode\vadjust pre{\hypertarget{ref-gannonShadowDeterrenceWhy2022}{}}%
Gannon, J. Andrés, Erik Gartzke, Jon R. Lindsay, and Peter Schram. 2022.
{``The {Shadow} of {Deterrence}: {Why Capable Actors Engage} in
{Contests Short} of {War}.''} \emph{Journal of Conflict Resolution},
00220027231166345.

>>>>

\author{
    Rex W. Douglass
    \thanks{Correspondence should be addressed to Rex W. Douglass at
\href{mailto:rexdouglass@gmail.com}{\nolinkurl{rexdouglass@gmail.com}}.}
   \\
    University of California, San Diego \\
   \\
  \texttt{} \\
   \And
    Thomas Leo Scherer
   \\
    University of California, San Diego \\
   \\
  \texttt{} \\
   \And
    J Andrés Gannon
   \\
    Vanderbilt University \\
   \\
  \texttt{} \\
   \And
    Erik Gartzke
   \\
    University of California, San Diego \\
   \\
  \texttt{} \\
   \And
    Jon Lindsay
   \\
    Georgia Institute of Technology \\
   \\
  \texttt{} \\
   \And
    Shannon Carcelli
   \\
    University of Maryland \\
   \\
  \texttt{} \\
   \And
    Jonathan Wilkenfeld
   \\
    University of Maryland \\
   \\
  \texttt{} \\
   \And
    David M. Quinn
   \\
    University of Maryland \\
   \\
  \texttt{} \\
   \And
    Catherine Aiken
   \\
    Georgetown University \\
   \\
  \texttt{} \\
   \And
    Jose Miguel Cabezas Navarro
   \\
    Universidad Mayor \\
   \\
  \texttt{} \\
   \And
    Neil Lund
   \\
    University of Maryland \\
   \\
  \texttt{} \\
   \And
    Egle Murauskaite
   \\
    University of Maryland \\
   \\
  \texttt{} \\
   \And
    Diana Partridge
   \\
    University of Maryland \\
   \\
  \texttt{} \\
  }

@@

Much like the recording of chess evolved from natural language
descriptive text notation to the modern figurine algebra notion,
international relations scholars have recently sought to produce a
structured account of historical events by combining the unstructured
corpus of historical records with informative priors about international
relations. The resulting structured account of information about a
historical episode can be combined with that of other events to produce
a systematic account of political events from which we can garner novel
insight, assuming the underlying data have high coverage, precision, and
recall. The easiest way to convey the desired produced of this task is
with an example. Figure 1 shows a narrative account of the Cuban Missile
Crisis (1962) in natural language sentences alongside a mapping to
discrete machine-readable abstractive events. From the structured data,
scholars can identify similarities and differences across events
concerning important concepts like when particular foreign policy
actions deter versus inflame (Jervis 1978; Glaser 2000), when third
parties mediate in interstate disputes (Haffar 2002; Quinn et al. 2006),
and how actors try to communicate resolve (Trager 2016; Lupton 2018).
Identifying patterns of international interactions is not just an
inherently interesting enterprise; it is a necessary precondition to
important efforts to predict where policymakers should turn their
attention to improve global welfare (Ward et al. 2013; Beger, Morgan,
and Ward 2021).

\begin{figure}
\hypertarget{fig-cubanarrative}{%
\centering
\includegraphics[width=7in,height=8in]{case_study_cuban_precision.png}
\caption{Cuban Missile Crisis (1962) - ICB Narrative vs.~ICBe
Events}\label{fig-cubanarrative}
}
\end{figure}

@@

With the exception of large-scale CAMEO dictionary-based systems (the
first grouping of columns), our ontology improves upon the the existing
state of the art quantitative datasets that ignore important information
content about international interactions.\footnote{See Balali, Asadpour,
  and Jafari (2021) for a recent review of ontological depth and
  availability of Gold Standard example text.} We highlight two
particular innovations. First, we separate the `chess pieces' from the
`chess players' in distinguishing between different actors within a
state. By virtue of our ontology coding military versus civilian actors
and national leaders versus bureaucrats, our data can be used to explore
important questions concerning civilian-military relations (Narang and
Talmadge 2017), Track Two diplomacy and the role of sub-national actors
(Hsu et al. 2020), and the evolution of what actors are engaged in
crises - a topic of increasing interest as states engage in gray zone
conflict by employing the coast guard or paramilitary mercenaries
instead of internationally recognized state militaries (Gannon et al.
2022). Second, we add information about the domains in which actors
behave - whether in land, air, sea, space, or cyber - since they differ
in their technology, tactics, geography, and purpose (Gartzke and
Lindsay 2019). Doing so allows researchers to identify and explain
patterns in escalation conditional on the military means states use in
conflict. Recent concerns about cross-domain conflict and the effect of
new domains of conflict like space and cyber have made this an endeavor
of increased interest to practitioners (Gannon 2022).

@@

The ICBe ontology follows a hierarchical design philosophy where a
smaller number of significant decisions are made early on and then
progressively refined into more specific details (Brust and Denzler
2020).\footnote{This process quickly focuses the coder on a smaller
  number of relevant options while also allowing them to apply multiple
  tags if the sentence explicitly includes more than one or there is
  insufficient evidence to choose only one tag. The guided coding
  process also allows for the possibility that earlier coarse decisions
  have less error than later fine-grained decisions.} Each coder was
instructed to first thoroughly read the full crisis narrative and then
presented with a custom graphical user interface (GUI) (SI Appendix
2.1). Coders then proceeded sentence by sentence, choosing the number of
events (0-3) that occurred, the highest behavior (thought, speech, or
activity), a set of players, whether the means were primarily armed or
unarmed, whether there was an increase or decrease in aggression
(uncooperative/escalating or cooperative/de-escalating), and finally one
or more specific and non-mutually exclusive activities. Some additional
details were always collected (e.g.~location and timing) while other
details were only collected if appropriate (e.g.~force size, fatalities,
domains, units). We find 472 actors and 117 different
behaviors.\footnote{See the full codebook on Github Repository
  \href{https://urldefense.com/v3/__https://github.com/CenterForPeaceAndSecurityStudies/ICBEventData__;!!Mih3wA!WxDJtEczKfxGTh0S2Krunap8ReymFEL5iTWaSfOHeqlSdyfRx77zmjBSWO1OAm13$}{ICBEventData}.}

A unique feature of the ontology is that thought, speech, and do
behaviors can be nested into combinations, e.g.~an offer for the
U.S.S.R. to remove missiles from Cuba in exchange for the U.S. removing
missiles from Turkey. Through compounding, the ontology can capture what
players were said to have known, learned, or said about other specific
fully described actions. Two other novel features of our data -- its
re-programmamable coding and its inclusion of thought, speech, and
action -- intersect to provide a unique contribution. No existing event
data distinguishes thoughts, speeches, and actions and most are limited
to just coding actions given they are the most readily observable and
thus the full universe could at least theoretically be
included.\footnote{Even the coding of overt actions like MIDs is not
  without contention (Gibler 2018).} Yet, methods like archival work and
expert interviews continue to improve identification of thoughts and
speeches that help us understand the reasoning process underlying
political decisions and what actually occurred behind the scenes. This
is an improvement over the state of the art that omits these by design
in allowing scholars to identify thoughts associated with things like
crisis escalation as well as patterns of omission to be advanced in
future work, but the modular nature of our coding means the inclusion of
a non-random sample of thoughts and speeches is neither necessary nor
does it have inevitable downstream effects on the rest of the coding in
ways that produce unwanted bias.

@@

Our strategy for dealing with both problems is a plausibly objective
yardstick called a synthetic historical narrative. We collect a large
diverse corpus of narratives spanning timelines, encyclopedia entries,
journal articles, news reports, websites, and government documents.
Using natural language processing (fully described in SI Appendix 3.1),
we identify details that appear across multiple accounts. The more
accounts that mention a detail, the more central it is to understanding
the true historical episode. The theoretical motivation is that authors
face word limits which force them to pick and choose which details to
include, and they choose details that serve the specific context of the
document they are producing. With a sufficiently large and diverse
corpus of documents, we can vary the context while holding the overall
episode constant and see which details tend to be invariant to context.
This presents a harder evaluation baseline than comparing ICBe's recall
to just that of ICB since there are non-crisis aspects of these events
that may be included in other narratives but are out of the scope of our
data. For example, the nationalization of businesses in Cuba may be
included as important context in the Cuban Missile Crisis in documents
that do not focus on the crisis dimensions like ICB. Using this hard
case, a recall measure of ICBe on the synthetic narratives thus serves
as a way to evaluate the breadth of ICBe's ontology and potential
application to non-crisis international events.

We find substantive variation in recall across existing state of the art
methods. Mentions of a detail across accounts are exponentially
distributed with context-invariant details appearing dozens to hundreds
of times more than context-dependent details.\footnote{As the ICB
  narratives are intended to explain conflictual behavior in a political
  context, many of the missing events concern more economic components
  of conflict (eg. nationalizing a foreign business). Even when they
  occur in the context of a crisis, these events largely fall outside
  the sample of information on which ICBe's ontology is currently
  trained. Even with this limitation, ICBe is more comprehensive than
  the existing datasets that do try to code the economic dimensions of
  these crises. We see expanding the ontology to broader international
  phenomenon as a promising future implementation of our model.}
Furthermore, crisis start and stop dates are arbitrary, and the
historical record points to many precursor events as necessary detail
for understanding later events. Figure 2 compares ICBe's recall with
that of existing datasets for the two case studies detailed in Section
4. ICBe strictly dominates all of the systems but ICEWs in recall though
we note that the small sample sizes mean these systems should be
considered statistically indistinguishable. Across all existing datasets
and ICBe, recall increases with the number of document mentions which is
an important sign of validity for both them and our benchmark. The one
outlier is Phoenix which in the Cuban Missile Crisis case is so noisy
that it's recall curve is flat to decreasing as mentions increase. The
two episode-level datasets (MIDs and ICM) have low coverage of
contextual details. The two other dictionary systems ICEWs and Terrier
have higher coverage, with ICEWs outperforming Terrier. Importantly our
corpus of ICB narratives has high recall of frequently mentioned details
giving us confidence in how those summaries were constructed, and ICBe
lags only slightly behind showing that it left little additional
information on the table.

@@

How does ICBe's precision compare to the existing state of the art? A
researcher should be able to lay out the events of a crisis on a
timeline and read off the macrostructure of an episode from each
individual move. We call this visualization a crisis map, a directed
graph intersected with a timeline. A crisis map using ICBe for the Cuban
Missile Crisis case study is provided in Figure 2, and crisis maps for
the two case studies using existing event datasets can be found in SI
Appendix 4.3 and 4.4 and crisis maps for all crises using all datasets
can be found on the companion website. The crisis maps reveal the
episode-level datasets like MIDs or the original ICB are too sparse and
vague to reconstruct the structure of the crisis (SI Appendix 4.3 and
4.4). On the other end of the spectrum, the high recall dictionary-based
event datasets like Terrier and ICEWs produce so many noisy events
(several hundred thousand) that even with heavy filtering their crisis
maps are completely unintelligible. Further, because of copyright
issues, none of these datasets directly provide the original text spans
making event-level precision difficult to verify.

@@

Figure 3 shows the location of every sentence from the ICBe corpus in
semantic space as embedded using the same large language model as
before, and the median location of each ICBe event tag applied to those
sentences.\footnote{We preprocess sentences to replace named entities
  with a generic Entity token.} Labels reflect the individual leaves of
the ontology and colors reflect the higher level coarse branch nodes of
the ontology. If ICBe has high precision, substantively similar tags
ought to have been applied to substantively similar source text, which
is what we see both in two dimensions in the main plot and via
hierarchical clustering on all dimensions in the dendrogram along the
right-hand side.\footnote{Hierarchical clustering on cosine similarity
  and with Ward's method.}

@@

In this section, we focus our validation on two case studies for which
we have produced synthetic narratives using the method described in
Section 3.2. Our proposed measure is a reconstruction task to see
whether our intended ontology can be recovered through only unsupervised
clustering of sentences they were applied to. The first (Figure 1) is
the Cuban Missile Crisis (hereafter Cuban Missiles) which took place
primarily in the second half of 1962, involved the United States, the
Soviet Union, and Cuba, and is widely known for bringing the world to
the brink of nuclear war. The second (SI Appendix 4.1) is the
Crimea-Donbas Crisis (hereafter Crimea-Donbas) which took place
primarily in 2014, involved Russia, Ukraine, and NATO, and within a
decade spiraled into a full-scale invasion. We choose these cases
because they are significant in contemporary international relations,
are widely known across academic disciplines as well as among the
public, and are sufficiently brief to evaluate in depth. They are
similar in that both cases involve a superpower in crisis with a
neighbor that changed from a friendly to a hostile regime, both held
implications for the economic and military security for the superpower
by risking full-scale invasion, and both eventually invited intervention
by an opposing superpower.

@@

A synthetic historical narrative for Cuban Missiles appears in Figure 4,
with 51 events drawn from 2,020 documents. Each row represents a detail
that appeared in at least five documents along with an approximate start
date, a handwritten summary, the number of documents it was mentioned
in, and whether it could be identified in the text of the original ICB
corpus, our ICBe events, and any of the competing existing models.

ICB's improved recall of Cuban Missiles was relative to the state of the
art was summarized in Section 3.2 (Figure 2), but the events that
explain that improvement can now be seen. Our ground truth ICB narrative
contains 17/51 of the events from the synthetic narrative of a case that
includes high-level previously classified details. ICBe captures nearly
all details included in ICB as well as more details from the synthetic
narrative than any competing dataset. Phoenix includes some earlier
information than ICBe like the nationalization of businesses and back
channel negotiations, but the crisis narrative has a clean canonical end
with the Soviets agreeing to withdraw missiles. ICBe stands out in
including more communicative behavior (do -- speech) than existing
datasets like US threats to attack and later promises not to invade.
Given the recognized importance of threat credibility for understanding
international conflict, the addition of this information is a
substantively important improvement over the existing state of the art
(Slantchev 2011).

@@

Figure 5 shows the crisis map for the Cuban Missile Crisis. Looking at
the crisis on a timeline, one can now identify the structure of actors
and the environment, along with its supporting details, in a way that
validates the precision of ICBe. Although harder to measure objectively,
this crisis map provides face valid evidence that ICBe's account is not
too vague, but also not unnecessarily detailed. We include much of the
geopolitically important details like Soviet deployment, US discovery of
that deployment, heightened alert levels, a blockade, and negotiations
that ended with a formal agreement. At the same time, the crisis map
indicates that ICBe does not include unnecessary nuances that preclude
useful comparison to other international events.

@@

A synthetic historical narrative for the 2014 Crimea-Donbas crisis (30
events drawn from 971 documents) appears in Figure 6. As in the earlier
case, rows represent details that appeared in at least five documents
and whether it is identified in ICBe and existing datasets.

Again quantitatively summarized earlier in Section 3.2 (Figure 2), our
ground truth ICB narrative contains 23/30 of the events from the
synthetic narrative. Like the gray zone precursor to the Cuban Missile
crisis (Cormac and Aldrich 2018), Ukraine provided several security
guarantees to Russia that were potentially undone, e.g.~a long term
lease on naval facilities in Crimea. But unlike the Cuban Missile
crisis, the end of this crisis is unclear, with the event meekly ending
with a second cease-fire agreement (Minsk II) but continued fighting.
ICBe again recalls more important information about the crisis than any
existing dataset, particularly information concerning the behavior of
non-state separatist groups like the Donetsk People's Republic (DPR) and
Luhansk People's Republic (LPR).

@@

We take the same approach here in Figure 7, selecting four specific
CAMEO event codings and checking how often they reflect a true
real-world event from the Crimea-Donbas synthetic narrative. We choose
four event types around key moments in the crisis. The start of the
crisis revolves around Ukraine backing out of a trade deal with the EU
in favor of Russia, but ``sign formal agreement'' events act more like a
topic detector with dozens of events generated by discussions of a
possible agreement but not the actual agreement which never
materialized. The switch is caught by the ``reject plan, agreement to
settle dispute'', but also continues for Viktor Yanukovych even after he
was removed from power because of articles retroactively discussing the
cause of his removal. Events for ``use conventional military force''
capture a threshold around the start of hostilities and who the
participants were but not any particular battles or campaigns. Likewise,
``impose embargo, boycott, or sanctions'' captures the start of waves of
sanctions and from who but are effectively constant as the news coverage
does not distinguish between subtle changes or additions. In sum,
dictionary-based methods on news corpora tend to have high recall
because they parse everything in the news, but for the same reason,
their specificity for most event types is too low to back out individual
chess-like sequencing that ICBe aims to record.

@@

We release several open-source products along with supporting code and
documentation to further advance the study of international relations,
event extraction, and natural language processing. The first is the
International Crisis Behavior Events (ICBe) dataset, an event-level
aggregation of what took place during the crises identified by the ICB
project. These data are appropriate for statistical analysis of hard
questions about the sequencing of events (e.g.~escalation and
de-escalation of conflicts).\footnote{Using ICBe data, Gannon (2022)
  finds that cross-domain crises are shorter and less violent than
  same-domain crises.} Second, we provide a coder-level disaggregation
with multiple codings of each sentence by experts and undergrads that
allows for the introduction of uncertainty and human interpretation of
events. Further, we release a direct mapping from the codings to the
source text at the sentence level as a new resource for natural language
processing. Finally, we provide a companion website that incorporates
detailed visualizations of all of the data introduced here at
crisisevents.org.

@@

\leavevmode\vadjust pre{\hypertarget{ref-gannonShadowDeterrenceWhy2022}{}}%
Gannon, J Andrés, Erik Gartzke, Jon R. Lindsay, and Peter Schram. 2022.
{``The {Shadow} of {Deterrence}: {Why Capable Actors Engage} in
{Contests Short} of {War}.''} \emph{Journal of Conflict Resolution},
00220027231166345.
